Presheaf Model for Dependent Types

### 2017-09-21 papers/talk-11/paper-11.tex
Improve formatting of the proofs and refer to STLC

As a motivation for dependent products recall how to define exponents in
presheaf categories. Let $\Gamma$ and $\Delta$ be presheaves and suppose we
already know how the exponent $\Delta^\Gamma$ is constructed. Then we get by
the Yoneda Lemma and the fact that $-^\Gamma$ should be right adjoint to $-
\times \Gamma$, that
\[
  (\Delta^\Gamma)(I) \cong Hom(\mathbf{y}I, \Delta^\Gamma) \cong Hom(\mathbf{y}I \times \Gamma, \Delta)
\]
where $\mathbf{y}$ is Yoneda embedding.

The latter can be taken up as a definition. So an element $w$ of
$(\Delta^\Gamma)(I)$ is a natural transformation $w \from \mathbf{y}I\times
\Gamma \to \Delta$, so it is given by functions
\[
  w_J \from (\mathbf{y}I)(J) \times \Gamma(J) \to \Delta(J),
\]
and the naturality condition becomes $(w_J(f, \rho))g = w_K(fg, \rho g)$ for $f
\from J \to I$, $\rho \in \Gamma(J)$, and $g \from K \to J$.

We will now show that CwF associated to a presheaf category supports
$\Pi$-types. Given $\Gamma \vdash A$ and $\Gamma.A \vdash B$ we have to define
$\Gamma \vdash \Pi x: A. B~type$, that is, we have to define the set $(\Pi x:
A.B) \rho$ for $I \in \cat{C}$ and $\rho \in \Gamma(I)$. The elements $w$ of
$(\Pi x:A.B)\rho$ are families $w = (w_f~|~J\in \cat{C}, f \from J \to I)$ of
(dependent) functions such that
\[
  w_f u \in B(\rho f, u) \qquad \text{for $J \in \cat{C}, f \from J \to I, and u \in A(\rho f)$},
\]
with the requirement that for $g \from K \to J$
\[
  (w_f u)g = w_{fg}(ug).
\]

@@

To define the application let $\Gamma \vdash u : \Pi x:A.B$ and $\Gamma \vdash
v : A$, and we set $App(u, v)\rho = (u\rho)_1 (v\rho) \in B(\rho, v\rho)$, thus
$App(u, v)\rho \in B(1, v)\rho$ as $(\rho, v \rho) = (1, v)\rho$.
$\beta$-equality is readily checked
\[
  App(\lambda b, v)\rho = ((\lambda b)\rho)_1(v\rho)=b(\rho, v\rho)
\]
and similarly for $\eta$-equality and other equations.

@@

\subsection*{Dependent sum}
The interpretation of $\Gamma \vdash \Sigma x:A.B~type$ for $\Gamma \vdash
A~type$ and $\Gamma.A \vdash B~type$ is defined by
\[
  (\Sigma x :A. B)\rho = \{(a, b)~|~a \in A\rho \land b \in B(\rho, a)\}
\]
for $\rho \in \Gamma(I), I \in \ob(\cat{C})$. The restrictions are defined
componentwise $(a, b)f = (af, bf)$ for $f \from J \to I$. The pairing operation
$\Gamma \vdash (u, v) : \Sigma x:A.B$ of $\Gamma \vdash u : A$ and $\Gamma.A
\vdash v : B(1, u)$ is defined by the componentwise pairing: $(u, v)\rho =
(u\rho, v\rho)$. Likewise for the projections $\mathbf{p}$ and $\mathbf{q}$,
$(\mathbf{p}w)\rho = a$ and $(\mathbf{q}w)\rho = b$ for $w\rho = (a, b)$. This
validates the necessary equations.

>>>>

\begin{example}
  The type for the function $A \to A$ (where $A$ is a ground type) is expressed
  as an exponential $M(A)^{M(A)}$ in STLC with $M$ being the interpretation
  function and $M(A)$ an object corresponding to the ground type in cartesian
  closed category.
\end{example}

To represent the same type in the model with dependent types we will also use
exponentials, but on the presheaves. Let us consider how to define exponents in
presheaf categories. Let $\Gamma$ and $\Delta$ be presheaves and suppose we
already know how the exponent $\Delta^\Gamma$ is constructed. Then we get by
the Yoneda Lemma and the fact that $-^\Gamma$ should be right adjoint to $-
\times \Gamma$, that
\[
  (\Delta^\Gamma)(I) \cong Hom(\mathbf{y}I, \Delta^\Gamma) \cong Hom(\mathbf{y}I \times \Gamma, \Delta)
\]
where $\mathbf{y}$ is the Yoneda embedding.

We can take that as a definition. Then an element $w$ of $(\Delta^\Gamma)(I)$
is a natural transformation $w \from \mathbf{y}I\times \Gamma \to \Delta$ given
by functions
\[
  w_J \from (\mathbf{y}I)(J) \times \Gamma(J) \to \Delta(J),
\]
and the naturality condition becomes $(w_J(f, \rho))g = w_K(fg, \rho g)$ for $f
\from J \to I$, $\rho \in \Gamma(J)$, and $g \from K \to J$.

\begin{lemma}
  CwF associated to a presheaf category supports $\Pi$-types.
  \begin{proof}
    We give a proof sketch as seen in \cite{Huber}. More detailed proof may be
    found in \cite{Hofmann}.

    Given $\Gamma \vdash A$ and $\Gamma.A \vdash B$ we have to define $\Gamma
    \vdash \Pi x: A. B~type$, that is, we have to define the set $(\Pi x: A.B)
    \rho$ for $I \in \cat{C}$ and $\rho \in \Gamma(I)$. The elements $w$ of
    $(\Pi x:A.B)\rho$ are families $w = (w_f~|~J\in \cat{C}, f \from J \to I)$
    of (dependent) functions such that
    \[
      w_f u \in B(\rho f, u) \qquad \text{for $J \in \cat{C}, f \from J \to I, and u \in A(\rho f)$},
    \]
    with the requirement that for $g \from K \to J$
    \[
      (w_f u)g = w_{fg}(ug).
    \]

@@

    To define the application let $\Gamma \vdash u : \Pi x:A.B$ and $\Gamma
    \vdash v : A$, and we set $App(u, v)\rho = (u\rho)_1 (v\rho) \in B(\rho,
    v\rho)$, thus $App(u, v)\rho \in B(1, v)\rho$ as $(\rho, v \rho) = (1,
    v)\rho$. $\beta$-equality is readily checked
    \[
      App(\lambda b, v)\rho = ((\lambda b)\rho)_1(v\rho)=b(\rho, v\rho)
    \]
    and similarly for $\eta$-equality and other equations.
  \end{proof}
\end{lemma}

@@

\subsection*{Dependent sum}
The interpretation of $\Gamma \vdash \Sigma x:A.B~type$ for $\Gamma \vdash
A~type$ and $\Gamma.A \vdash B~type$ is defined by
\[
  (\Sigma x :A. B)\rho = \{(a, b)~|~a \in A\rho \land b \in B(\rho, a)\}
\]
for $\rho \in \Gamma(I), I \in \ob(\cat{C})$.
\begin{lemma}
  CwF associated to a presheaf category supports $\Sigma$-types.
  \begin{proof}
    Again we give a proof sketch as seen in \cite{Huber}. More detailed proof
    is also found in \cite{Hofmann}.

    The restrictions are defined componentwise:
    \[
      (a, b)f = (af, bf)
    \]
    for $f \from J \to I$.

    The pairing operation $\Gamma \vdash (u, v) : \Sigma x:A.B$ of $\Gamma
    \vdash u : A$ and $\Gamma.A \vdash v : B(1, u)$ is defined by the
    componentwise pairing:
    \[
      (u, v)\rho = (u\rho, v\rho)
    \]

    Likewise for the projections $\mathbf{p}$ and $\mathbf{q}$,
    $(\mathbf{p}w)\rho = a$ and $(\mathbf{q}w)\rho = b$ for $w\rho = (a, b)$.
    This validates the necessary equations.
  \end{proof}
\end{lemma}
